Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-12-18 agujournaltemplate.tex
working on reviewer #1 response

% MY OWN ADDED PACKAGES TO RENDER UNITS
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage[section]{placeins}

@@

\begin{keypoints}
\item The CARIACO phytoplankton community time series is characterized by two distinct temporal clusters separated by a shift in 2004 from high to low upwelling and by a return to high upwelling conditions in 2014.
\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in local weather and upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the Bermuda Atlantic Time-series Study (BATS) and Hawaii Ocean Time-series (HOT) sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.

@@

    Between late 1995 and early 1998, nutrient data, including nitrate, phosphate, and silicate, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr. K. Fanning (University of South Florida, St. Petersburg, USA). Samples from eight cruises were analyzed by both laboratories \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.

@@

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size). For simplicity, we refer to this variable as ``wind speed'' throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}. In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores, $z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means. Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Analysis of Phytoplankton Community Structure}
    To examine temporal changes in the phytoplankton community structure in the Cariaco Basin, we applied complete linkage hierarchical clustering using the ``hclust'' function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix This analysis was based on presence-absence data of observed genera, over the upper 100 meters of the water column and aggregated annually. By using binary data and temporal aggregation, we aimed to reduce the influence of extreme values in the cell counts and focus on broad compositional shifts in the community. 
    Additionally, we performed non-metric multidimensional scaling (NMDS) ordination on the monthly counts, depth-integrated to \qty{100}{\meter}, that were square-root transformed using ``metaMDS'' from the ``vegan'' package \cite{oksanen_vegan_2024}. The NMDS analysis is based on the Bray distance matrix with a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the ``envfit'' function from the ``vegan'' package to fit the environmental vectors onto the ordination with 999 permutations. In contrast to the annual clustering, the NMDS ordination uses the individual monthly cell counts and takes abundance of genera into account.
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the ``ggpubr'' package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the ``wilcox.test'' function in R \cite{r_core_team_r_2024}.
    % detailed explanation of NMDS in Winder and Hunter 2008

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, in log-transformed chlorophyll \textit{a}, and in phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift in the community. This shift persists during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years when mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

@@

    % Figure 3
    Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts} $a$). The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards. Nevertheless, the monthly chlorophyll \textit{a} measurements for the remainder of the year 2013 were markedly lower (65\,\% of the mean over the entire time series for these months), which recovered in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter. Genus richness was consistently highest in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series. Across the time series, we observed maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, followed by a recovery after 2011. The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts} $a$). Scaling genus richness by the cell count abundances, i.e. the Shannon index, showed a similar but less pronounced dynamic. Diversity remained high throughout 2000 and recovered more quickly but showed marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal was lost (see Figure \ref{fig:divts} $f$).

    Community evenness, expressed by the Pielou index, shows no seasonal signal and a very large variance within seasons and through depth (see Figure \ref{fig:divts} $h$). The inter-annual dynamics of evenness are remarkably stable across all depth intervals, clustered around an intermediate 0.5 value. There were marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009, and 2016. Throughout 2003 to 2013 evenness was almost consistently highest in the deepest interval below 75 meters, where the community appeared to have been more evenly distributed.

@@

\subsection{AMO and upwelling-related variables drive community changes}
    % Figure 6
    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF} $a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community. Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high. For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera). 
    Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. Variables that show a pronounced threshold response are evaporation and salinity. For evaporation the threshold is at larger values of evaporation, which are most likely linked to periods of strong stratification, and salinity at both extreme ends of the range, signifying effects on cell counts of diatoms, dinoflagellates and cyanobacteria at strong upwelling and stratified conditions.

@@

\subsection{Community reorganization and biomass shifts}
    % P1: Introductory paragraph highlighting the overall dynamics/shifts
    This study shows the long-term dynamics of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
    These changes are clearly visible in the yearly mean anomalies and the density distribution of monthly measurements. After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014. 
    By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations. Concurrently, we observed an increase in chlorophyll \textit{a} and a return to a community composition more similar to the period between 1996 and 2004.

@@

    % Disucssion of HPLC vs Microscopy!
    A similar general pattern emerged when comparing our results with those based on High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}. However, some differences emerged between our results and those obtained with the HPLC data. The HPLC data showed an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. Total chlorophyll \textit{a} could not be quantified with a consistent method across the different laboratories \cite{pinckney_phytoplankton_2015}. This is the reason why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
    Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
    Additionally, despite decreased cell counts of all functional groupsing, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}. This provides a further strong indication that the phytoplankton community shifted to smaller cell sizes, which were not detected using light microscopy.

>>>>

% MY OWN ADDED PACKAGES TO RENDER UNITS
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{siunitx}
\sisetup{separate-uncertainty=true}
\usepackage[section]{placeins}

@@

\begin{keypoints}
\item The CARIACO phytoplankton time series reveals two temporal clusters, linked to a 2004 shift from high to low upwelling and return in 2014.
\item Diversity metrics show no significant difference except lower turnover and higher evenness during the cluster with low upwelling conditions.
\item Gradient forest analysis finds AMO index as strongest predictor of community shifts, highlighting large-scale climate impacts on ecosystem.
\end{keypoints}

@@

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to variations in local weather and upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned a state similar to that before the shift in 2004. In contrast, diversity declined between 1998 and 2009 independently of changes in the upwelling regime.
To quantify effects of climate and bottom-up drivers on phytoplankton community changes, we employed gradient forest analysis and found the AMO index was the strongest predictor, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive (mean integrated primary production across the entire time series of \qty{485 \pm 345}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the Bermuda Atlantic Time-series Study (BATS) and Hawaii Ocean Time-series (HOT) sites. This high productivity, combined with reduced ventilation across the sill, resulted in the formation of anoxic conditions below a depth of \qty{\sim 275}{m}.

@@

    Between late 1995 and early 1998, nutrient data, including nitrate, phosphate, and silicate, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr. K. Fanning (University of South Florida, St. Petersburg, USA). There is an overlap of eight nutrient samples analyzed by both laboratories \cite{taylor_ecosystem_2012}. We used a merged dataset for all nutrients, in which these eight coincidental monthly measurements are averaged.

@@

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the CARIACO time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size). For simplicity, we refer to this variable as ``wind speed'' throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}. In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores, $z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means. For the analysis of annual z-scores, data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Analysis of Phytoplankton Community Structure}
    To examine temporal changes in the phytoplankton community structure in the Cariaco Basin, we applied complete linkage hierarchical clustering using the ``hclust'' function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix. This analysis was based on presence-absence data of observed genera, over the upper 100 meters of the water column and aggregated annually. By using binary data and temporal aggregation, we aimed to reduce the influence of extreme values in the cell counts and focus on broad compositional shifts in the community. 
    Additionally, we performed non-metric multidimensional scaling (NMDS) ordination on the monthly counts, depth-integrated to \qty{100}{\meter}, that were square-root transformed using ``metaMDS'' from the ``vegan'' package \cite{oksanen_vegan_2024}. The NMDS analysis is based on the Bray distance matrix with a two-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the ``envfit'' function from the ``vegan'' package to fit the environmental vectors onto the ordination with 999 permutations. In contrast to the annual clustering, the NMDS ordination uses the individual monthly cell counts and takes abundance of genera into account.
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the ``ggpubr'' package \cite{kassambara_ggpubr_2023}. To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the ``wilcox.test'' function in R \cite{r_core_team_r_2024}.
    % detailed explanation of NMDS in Winder and Hunter 2008

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, in log-transformed chlorophyll \textit{a}, and in phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate. Instead, annual anomalies show an inverse relationship with precipitation and salinity, indicating effects of surface runoff. Elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift in the community. 
    This biological shift was preceded by a change in the AMO index anomalies that occurred after 2002, when the index transitioned to a period of mostly positive anomalies that persisted until 2013. This period was also characterized by upwelling-related variables indicating a weak upwelling regime (low wind, high SST, deep isotherm). Thus, while the AMO and physical parameters shifted after 2002, the corresponding change in chlorophyll a and cell counts occurred later, after 2004.
    Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity z-scores are elevated between 2007 and 2013, when chlorophyll \textit{a} and cell counts were diminished. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

@@

    % Figure 3
    Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 1999, 2005, 2008, 2009, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts} $a$). The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards. Nevertheless, the monthly chlorophyll \textit{a} measurements for the remainder of the year 2013 were markedly lower (65\,\% of the mean over the entire time series for these months), which recovered in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. During the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons. Genus richness was consistently highest in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series. Across the time series, we observed maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, followed by a recovery after 2011. The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts} $a$). Scaling genus richness by the cell count abundances, i.e. the Shannon index, showed a similar but less pronounced dynamic. Diversity remained high throughout 2000 and recovered more quickly but showed marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016. The seasonal signal that is apparent for chlorophyll \textit{a} and genus richness is not visible in the Shannon index (see Figure \ref{fig:divts} $f$).

    Community evenness, expressed by the Pielou index, shows no seasonal signal and a very large variance within seasons and through depth (see Figure \ref{fig:divts} $h$). The inter-annual dynamics of evenness were remarkably stable across all depth intervals, clustered around an intermediate 0.5 value. There were marked reductions in all depths towards 2000 and for all depths above 75 meters in 2007, 2009, and 2015. Throughout 2003 to 2013 evenness was almost consistently highest in the deepest interval below 75 meters, where the community appeared to have been more evenly distributed.

@@

\subsection{AMO and upwelling-related variables drive community changes}
    % Figure 6
    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF} $a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community. 
    Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high. For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera). 
    Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black lines indicate cumulative importance across the entire community, and the colored lines indicate importance per phytoplankton group. 
    Cumulative importance curves from gradient forest analysis show how the importance of a predictor accumulates across its range of values. The slope of the curve at any point indicates where the strongest compositional turnover occurs—steep sections represent threshold values where small changes in the predictor correspond to large changes in community composition, while flat sections indicate regions where the predictor has little influence on community structure.
    For the AMO index, we see a clear pattern in which negative anomalies affect the community (slope is highest for negative values). This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. Variables that show a pronounced threshold response are evaporation and salinity. For evaporation the threshold is at larger values of evaporation, which are most likely linked to periods of strong stratification, and salinity at both extreme ends of the range, signifying effects on cell counts of diatoms, dinoflagellates and cyanobacteria at strong upwelling and stratified conditions.

@@

\subsection{Community reorganization and biomass shifts}
    % P1: Introductory paragraph highlighting the overall dynamics/shifts
    This study shows the long-term dynamics of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes. In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
    These changes are clearly visible in the yearly mean anomalies and the density distribution of monthly measurements. After 2004, we observed increased SST, decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the AMO index toward positive mean anomalies until 2014. 
    By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}. In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations. Concurrently, we observed an increase in chlorophyll \textit{a} and a return to a community composition more similar to the period between 1996 and 2004.

@@

    % Disucssion of HPLC vs Microscopy!
    A similar general pattern emerged when comparing our results with those based on High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}. However, some differences emerged between our results and those obtained with the HPLC data. The HPLC data showed an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}. It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006. Total chlorophyll \textit{a} could not be quantified with a consistent method across the different laboratories \cite{pinckney_phytoplankton_2015}. This is the reason why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
    Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
    Additionally, despite decreased cell counts of all functional grouping, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}. This provides a further strong indication that the phytoplankton community shifted to smaller cell sizes, which were not detected using light microscopy.
